---
title: "Change-of-support field smoothing with gpfield"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Change-of-support field smoothing with gpfield}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5
)
set.seed(1L)
```

```{r libs}
library(gpfield)
```

## What this vignette shows

`gpfield` fits a spatial Gaussian process to a field or trial layout and answers
three questions a noisy plot grid alone cannot:

1. what is the smoothed trait surface, with calibrated uncertainty;
2. what is the trait averaged over a paddock-scale block -- a *change of support*
   from the observed point process to an areal one, with the predictive variance
   propagated through the support map rather than averaged away;
3. when is the field simply too coarse, or a block too far from any data, for an
   honest answer -- in which case `gpfield` abstains rather than fabricate a
   confident surface.

We use a real, public multi-environment trial -- the *Besag MET* corn-yield
trial distributed in the `agridat` package -- so every figure below comes from
data you can load yourself. The trial is documented in `agridat` and originates
with Besag and Higdon (1999).

## A real row-column trial

The trial is laid out as a row-by-column grid of plots within each of six
environments (`county`), each an 18-by-11 lattice of 64 corn hybrids. We take
one environment and treat its plot grid as the spatial field. A handful of
plots carry a missing yield -- `gpfield` drops incomplete rows before fitting,
and the count below confirms how many.

```{r data}
if (!requireNamespace("agridat", quietly = TRUE)) {
  knitr::knit_exit()
}
met <- agridat::besag.met
field <- met[met$county == "C1", c("row", "col", "yield")]
str(field)
sum(is.na(field$yield))
```

## Fitting the Gaussian process

A model is declared with `gpfield_spec()` -- which covariance kernel, which
coordinate columns, which response -- and fitted with `gp_fit()`. The default is
a `Matern` covariance at smoothness `1.5`, the usual choice for field data: it is
once-differentiable rather than the squared-exponential's infinite smoothness,
which avoids over-smoothing real agronomic gradients.

```{r fit}
spec <- gpfield_spec(coords = c("row", "col"), response = "yield")
fit <- gp_fit(spec, field, seed = 1L)
fit
```

The fit reports an estimated correlation `range` in raw plot units. That range is
the quantity the rest of the package reasons about: a prediction spaced more
coarsely than the range asks the GP to interpolate between effectively
independent observations and is unsupported, and the fit makes that explicit
instead of leaving it implicit. Refining a prediction grid (see below) makes it
*finer*, not coarser, so refining never by itself triggers this abstention --
only an observed grid coarser than the range does, at every refinement.

## Smoothing the surface

`gp_field_smooth()` is the one-call path from a row-column layout to a smoothed
surface. Here we refine the prediction grid to twice the plot resolution and draw
the result (the plot needs `ggplot2`).

```{r smooth, fig.alt = "Smoothed corn-yield surface over the row-column plot grid", fig.cap = "Posterior mean corn-yield surface for county C1, smoothed onto a grid twice as fine as the plot layout (refine = 2). Warmer cells are higher predicted yield."}
smoothed <- gp_field_smooth(field, response = "yield", refine = 2L, seed = 1L)
smoothed@prediction

if (requireNamespace("ggplot2", quietly = TRUE)) {
  gp_field_plot(smoothed)
}
```

The surface rises toward the high-row, high-column corner of the lattice --
both axes carry a positive trend in the raw plots (`cor(row, yield)` $\approx
0.50$, `cor(col, yield)` $\approx 0.20$) -- rather than showing the noisy
plot-to-plot scatter of the raw data; that smoothing is the Matern range doing
its job of separating spatial signal from plot-level noise. Refining the grid
to twice the plot resolution does not invent new information, only a finer
canvas on which the same fitted surface is drawn, which is why the abstention
rule above binds on the *observed* grid's spacing, not the refined one.

## Change of support: from plots to blocks

The defining capability is predicting at a *different support*: reconciling
observations and predictions made at different spatial scales, the general
change-of-support problem set out by Cressie (1993) and surveyed by Gotway and
Young (2002). A point process (the plots) can be aggregated to an areal one
(paddock-scale blocks). `gpfield` approximates the block average by a
tensor-midpoint quadrature rule over $n_q$ nodes inside the block footprint,
with equal weights $w_j = 1/n_q$. Writing $K_{*n}$ for the covariance between
the $n_q$ quadrature nodes and the $n$ training points, $K_n$ for the training
covariance, and $K_{**}$ for the quadrature nodes' own covariance, the block
mean and variance are

$$
\hat{y}_B = \sum_{j=1}^{n_q} w_j \, \big[K_{*n} K_n^{-1} y\big]_j, \qquad
\mathrm{Var}(y_B) = w^\top \big(K_{**} - K_{*n} K_n^{-1} K_{n*}\big) w.
$$

Both are the same posterior mean and covariance a point prediction would use,
evaluated at the quadrature nodes and then averaged with the quadrature
weights -- so the block variance carries the full within-block posterior
covariance, not just the marginal point-wise variances. That off-diagonal term
is what makes the block-average uncertainty larger than a naive average of
independent point predictions.

```{r blocks}
blocks <- block_support(
  lower = rbind(c(1, 1), c(10, 1)),
  upper = rbind(c(9, 11), c(18, 11)),
  ids = c("south", "north")
)
block_pred <- gp_predict(fit, support = "block", blocks = blocks, seed = 1L)
block_pred@locations
data.frame(block = block_pred@locations$block,
           mean = round(block_pred@mean, 1),
           sd = round(block_pred@sd, 2))
```

Why this is not a cell-wise average: a naive mean of `n` point predictions would
assume independence and shrink the block-average standard deviation by
`1 / sqrt(n)`. Because neighbouring locations within a block are positively
correlated, the change-of-support variance is materially larger -- the honest
uncertainty for a paddock-scale claim.

```{r cos-compare}
centre <- gp_predict(fit, data.frame(row = 5, col = 6))
# The tensor-midpoint rule realises per_axis^2 nodes, per_axis = round(sqrt(n_quad)),
# not the requested n_quad exactly -- divide by the realised count, not the request.
per_axis <- max(2L, round(sqrt(blocks$n_quad)))
n_realised <- per_axis^2L
naive_sd <- centre@sd / sqrt(n_realised)
c(change_of_support_sd = round(block_pred@sd[[1L]], 3),
  naive_independence_sd = round(naive_sd, 4))
```

## Honest abstention

`gpfield` refuses a prediction the data cannot support. Two refusals:

```{r abstain-coarse}
# A query grid spaced far more coarsely than the estimated range.
coarse <- gp_predict(fit, data.frame(row = c(1, 50, 100),
                                     col = c(1, 50, 100)))
is_gpfield_abstention(coarse)
coarse
```

```{r abstain-gap}
# A block placed far outside the trial footprint.
far <- block_support(lower = rbind(c(100, 100)), upper = rbind(c(110, 110)))
gp_predict(fit, support = "block", blocks = far, seed = 1L)
```

Each abstention carries a machine-readable `reason` and the numbers behind it, so
a downstream caller -- or the orchestration layer -- can branch on the refusal
rather than catch an exception or, worse, trust a confident wrong value.

## Emitting the orchestra contract

`gpfield` is an orchestra member, so a result is expressible in the federation's
shared manifest contract: a versioned, integrity-hashed object every member emits
or consumes. `as_orchestra_manifest()` adapts a prediction into it, and
`verify_manifest()` confirms the payload was not tampered with.

```{r manifest}
point_pred <- gp_predict(fit, field[, c("row", "col")])
manifest <- as_orchestra_manifest(point_pred)
manifest
verify_manifest(manifest)$ok
```

The smoothed surface is now expressed in the orchestra's shared manifest
contract, the format its calibration, uncertainty-quantification,
causal-inference and decision members are built to consume.

## Scaling to large fields

The exact solver is cubic in the number of observations -- comfortable for the
hundreds-to-low-thousands of plots a multi-environment trial carries. For a field
too large for the exact solver, `solver = "lowrank"` fits a Deterministic
Training Conditional (DTC) sparse Gaussian process (Quinonero-Candela and
Rasmussen 2005) on a set of $m$ space-filling inducing points, at cost linear in
the number of plots $n$ and quadratic in the (much smaller) inducing count.
Writing $K_{nm}$ for the covariance between training points and inducing
points and $K_{mm}$ for the inducing points' own covariance, DTC replaces the
exact $n \times n$ training covariance $K_n$ by the rank-$m$ approximation

$$
K_n \approx K_{nm} K_{mm}^{-1} K_{mn},
$$

so every quantity that would otherwise need an $n \times n$ Cholesky
factorisation instead needs only an $m \times m$ one, at cost
$O(n m^2)$ rather than $O(n^3)$. Every verb above -- point and block
prediction, change of support, honest abstention, manifest emission -- is
unchanged; only the backend differs.

```{r lowrank}
set.seed(1L)
big <- expand.grid(x = seq(0, 25, by = 0.5), y = seq(0, 25, by = 0.5))
big$yield <- sin(0.3 * big$x) + cos(0.25 * big$y) + rnorm(nrow(big), 0, 0.2)
nrow(big)

fit_lr <- gp_fit(
  gpfield_spec(c("x", "y"), "yield", solver = "lowrank", n_inducing = 120L),
  big, seed = 1L
)
fit_lr
```

When the inducing set covers the whole field the low-rank posterior is exactly
the exact GP, so the choice is a scale-versus-cost dial rather than a change of
model. The default backend stays exact.

## References

Besag, J. and Higdon, D. (1999). Bayesian analysis of agricultural field
experiments. *Journal of the Royal Statistical Society, Series B*, 61(4),
691--746. Trial distributed in the `agridat` package.

Cressie, N. (1993). *Statistics for Spatial Data*, revised edition. Wiley.

Gotway, C. A. and Young, L. J. (2002). Combining incompatible spatial data.
*Journal of the American Statistical Association*, 97(458), 632--648.

Quinonero-Candela, J. and Rasmussen, C. E. (2005). A unifying view of sparse
approximate Gaussian process regression. *Journal of Machine Learning
Research*, 6, 1939--1959.
