---
title: "From point-sampled plots to a paddock block mean"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From point-sampled plots to a paddock block mean}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~5 s under rmarkdown::render() with agridat and ggplot2
     installed; macOS arm64, R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
```

```{r library}
library(gpfield)
```

## Why

A trialist who has just walked an 18-by-11 corn multi-environment trial, plot
by plot, does not actually want any one plot's yield. The management question
sits one level up: *if all I have are point samples from these plots, what is
the average yield -- and how uncertain is that average -- over the whole
paddock-scale block I am about to manage as one unit, and can a grid this
coarse even answer that, or should it say so and stop?* That is a
change-of-support problem: reconciling an observation made at one spatial
support (the plot) with a prediction wanted at another (the block), and
propagating the plots' spatial correlation into the block's uncertainty
rather than averaging it away. This vignette answers it twice over -- once
where the data support an answer, and once where `gpfield` refuses to give
one -- on a real, public multi-environment trial, so every number below is
reproducible from data anyone can load.

## What

Five verbs carry the whole of this vignette. `gpfield_spec()` declares the
model: which covariance kernel, which coordinate columns, which response.
`gp_fit()` estimates its hyperparameters by marginal-likelihood optimisation
and returns a `gpfield_fit` whose estimated correlation range is made
explicit, because that range is what every guard below reasons about.
`block_support()` describes one or more paddock-scale blocks as
axis-aligned rectangles, and `gp_predict(..., support = "block")` is the
change-of-support verb itself: the posterior mean and the full within-block
predictive covariance, integrated by quadrature and reduced to a block
average and its standard deviation. `gp_field_smooth()` is the one-call
path from a row-column trial layout to a smoothed surface, and
`gp_field_plot()` draws it. `as_orchestra_manifest()` and
`verify_manifest()` express a prediction in, and check, the orchestra's
shared manifest contract.

## Do

### Engines available on this build

Every chunk below that touches real data needs `agridat`; the figure needs
`ggplot2`. Both are checked once, and a guarded chunk explains what is
skipped when either is absent, so the vignette still renders -- with less
to show -- on a build that carries neither.

```{r engines}
has_agridat <- requireNamespace("agridat", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r engines-skip, eval = !has_agridat, echo = FALSE, results = "asis"}
cat("agridat is not installed on this build, so the corn-trial sections",
    "below are skipped; the low-rank section at the end runs on a",
    "synthetic field and is unaffected.\n")
```

### A real row-column trial

We use the corn-hybrid Besag MET (multi-environment trial), distributed in
the `agridat` package and originating with Besag and Higdon (1999): 64 corn
hybrids across six counties in North Carolina, each county laid out as an
18-by-11 lattice of plots. One county (`C1`) is treated as the spatial
field.

```{r data, eval = has_agridat, echo = has_agridat}
met <- agridat::besag.met
field <- met[met$county == "C1", c("row", "col", "yield")]
n_missing <- sum(is.na(field$yield))
str(field)
```

```{r data-kable, echo = FALSE, eval = has_agridat}
facts <- data.frame(
  fact = c("plots (rows of the field table)", "row indices",
           "column indices", "plots with missing yield",
           "correlation, row with yield", "correlation, column with yield"),
  value = c(
    as.character(nrow(field)),
    paste0(min(field$row), " to ", max(field$row)),
    paste0(min(field$col), " to ", max(field$col)),
    as.character(n_missing),
    sprintf("%.2f", cor(field$row, field$yield, use = "complete.obs")),
    sprintf("%.2f", cor(field$col, field$yield, use = "complete.obs"))
  ),
  stringsAsFactors = FALSE
)
knitr::kable(
  facts, col.names = c("Fact", "Value"),
  caption = "County C1 of the Besag MET corn trial."
)
```

`gpfield` drops the incomplete rows before fitting; the count above is what
it drops, not a hand count.

### Fitting the Gaussian process

A model is declared with `gpfield_spec()` and fitted with `gp_fit()`. The
default is a Matern covariance at smoothness 1.5, the usual choice for
field data: once-differentiable rather than the squared-exponential's
infinite smoothness, which avoids over-smoothing real agronomic gradients.

```{r fit, eval = has_agridat, echo = has_agridat}
spec <- gpfield_spec(coords = c("row", "col"), response = "yield")
fit <- gp_fit(spec, field, seed = 1L)
fit
```

The fit reports an estimated correlation range in raw plot units -- the
quantity the rest of the package reasons about. A prediction spaced more
coarsely than this range asks the Gaussian process to interpolate between
effectively independent observations, and every abstention guard below is
stated in units of it.

### Two paddock blocks over the trial

`block_support()` describes each management block as axis-aligned lower and
upper bounds on the same row/column coordinates the fit was built from. We
split the field lengthways into a south and a north half.

```{r blocks, eval = has_agridat, echo = has_agridat}
blocks <- block_support(
  lower = rbind(c(1, 1), c(10, 1)),
  upper = rbind(c(9, 11), c(18, 11)),
  ids = c("south", "north")
)
blocks
```

### The smoothed surface, with the block footprints marked

`gp_field_smooth()` fits and predicts back onto the plot grid in one call;
here we refine the prediction grid to twice the plot resolution.
`gp_field_plot()` draws the smoothed surface, and the two block outlines
from above are drawn over it -- a figure a table of block means cannot
give, because it shows where each block's average sits in the field, not
just what the average is.

```{r smooth, eval = has_agridat, echo = has_agridat}
smoothed <- gp_field_smooth(field, response = "yield", refine = 2L, seed = 1L)
smoothed@prediction
```

```{r fig-surface, eval = has_agridat && has_ggplot2, echo = has_agridat && has_ggplot2, fig.cap = "Posterior mean corn yield over county C1 of the Besag MET, smoothed onto a grid twice as fine as the plot layout (refine = 2), with the south and north paddock blocks used below outlined in white. Warmer cells are higher predicted yield; agridat does not document a physical unit for this trial's yield column, so the colour scale carries Besag and Higdon's original, unlabelled scale."}
block_df <- data.frame(
  block = blocks$ids,
  xmin = blocks$lower[, 1L], xmax = blocks$upper[, 1L],
  ymin = blocks$lower[, 2L], ymax = blocks$upper[, 2L]
)
gp_field_plot(smoothed) +
  ggplot2::geom_rect(
    data = block_df,
    ggplot2::aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE, colour = "white", fill = NA, linewidth = 0.9
  ) +
  ggplot2::labs(x = "plot row index (of 18)",
               y = "plot column index (of 11)", fill = "smoothed yield")
```

```{r fig-surface-skip, eval = !(has_agridat && has_ggplot2), echo = FALSE, results = "asis"}
cat("The smoothed-surface figure needs both agridat and ggplot2; it is",
    "skipped when either is absent.\n")
```

The surface rises toward the high-row end of the lattice -- both axes
carry a positive trend in the raw plots, and the county-C1 facts table
above gives the correlation of each -- rather than showing the noisy
plot-to-plot scatter of the raw data; that smoothing is the Matern range
doing its job of separating spatial signal from plot-level noise.

### Block-average predictions: mean and standard deviation

The change-of-support problem itself was set out in general by Cressie
(1993) and surveyed by Gotway and Young (2002): a point process (the
plots) aggregated to an areal one (the blocks). `gpfield` approximates
each block average by tensor-midpoint quadrature over $n_q$ nodes inside
the block footprint, with equal weights $w_j = 1 / n_q$. Writing $K_{*n}$
for the covariance between the $n_q$ quadrature nodes and the $n$ training
points, $K_n$ for the training covariance, and $K_{**}$ for the quadrature
nodes' own covariance, the block mean and variance are

$$
\hat{y}_B = \sum_{j=1}^{n_q} w_j \, \big[K_{*n} K_n^{-1} y\big]_j, \qquad
\mathrm{Var}(y_B) = w^\top \big(K_{**} - K_{*n} K_n^{-1} K_{n*}\big) w.
$$

Both are the same posterior mean and covariance a point prediction would
use, evaluated at the quadrature nodes and then averaged with the
quadrature weights, so the block variance carries the full within-block
posterior covariance rather than the marginal point-wise variances alone.

```{r block-predict, eval = has_agridat, echo = has_agridat}
block_pred <- gp_predict(fit, support = "block", blocks = blocks, seed = 1L)
block_tab <- data.frame(
  block = block_pred@locations$block,
  mean = round(block_pred@mean, 1L),
  sd = round(block_pred@sd, 2L),
  n_support = block_pred@n_support
)
```

```{r block-predict-kable, echo = FALSE, eval = has_agridat}
knitr::kable(
  block_tab, col.names = c("Block", "Mean yield", "SD", "Supporting plots"),
  caption = paste0(
    "Block-average predictions for the south and north halves of county C1."
  )
)
```

### Why this is not a cell-wise average

A naive mean of the same $n_q$ point predictions would treat them as
independent and shrink the block-average standard deviation by
$1 / \sqrt{n_q}$. Because neighbouring plots within a block are positively
correlated, the change-of-support variance is materially larger.

```{r cos-compare, eval = has_agridat, echo = has_agridat}
centre <- gp_predict(fit, data.frame(row = 5, col = 6))
# The tensor-midpoint rule realises per_axis^2 nodes, per_axis =
# round(sqrt(n_quad)), not the requested n_quad exactly -- divide by the
# realised count, not the request.
per_axis <- max(2L, round(sqrt(blocks$n_quad)))
n_realised <- per_axis^2L
naive_sd <- centre@sd / sqrt(n_realised)
sd_compare <- data.frame(
  quantity = c("change-of-support SD (south block)",
               "naive independence SD (centre point / sqrt(n))"),
  sd = round(c(block_pred@sd[[1L]], naive_sd), 3L)
)
```

```{r cos-compare-kable, echo = FALSE, eval = has_agridat}
knitr::kable(
  sd_compare, col.names = c("Quantity", "SD"),
  caption = paste0(
    "Change-of-support standard deviation against the naive independence ",
    "assumption, both over ", n_realised, " quadrature nodes."
  )
)
```

### Abstention: when the data will not bear the answer

`gpfield` refuses a prediction the data cannot support, returning a typed
`gpfield_abstention` with a machine-readable reason rather than a
confident surface built on data too sparse or too coarse to carry it. Two
refusals below, each triggered for real by the fit above, are read against
their own numbers in the *Read* section.

```{r abstain-coarse, eval = has_agridat, echo = has_agridat}
# A query grid spaced far more coarsely than the estimated range.
coarse <- gp_predict(fit, data.frame(row = c(1, 50, 100),
                                     col = c(1, 50, 100)))
is_gpfield_abstention(coarse)
coarse
```

```{r abstain-gap, eval = has_agridat, echo = has_agridat}
# A block placed far outside the trial footprint.
far <- block_support(lower = rbind(c(100, 100)), upper = rbind(c(110, 110)))
far_pred <- gp_predict(fit, support = "block", blocks = far, seed = 1L)
is_gpfield_abstention(far_pred)
far_pred
```

### Emitting the orchestra contract

`gpfield` is an orchestra member: a prediction is expressible in the
federation's C2 v2 contract, the generic `orchestra_manifest` every current
and future orchestra member emits or consumes. `as_orchestra_manifest()`
adapts a prediction into it; `verify_manifest()` confirms the payload was
not tampered with.

```{r manifest, eval = has_agridat, echo = has_agridat}
point_pred <- gp_predict(fit, field[, c("row", "col")])
manifest <- as_orchestra_manifest(point_pred)
manifest
verify_manifest(manifest)$ok
```

### Scaling to large fields

The exact solver is cubic in the number of observations -- comfortable for
the hundreds-to-low-thousands of plots a multi-environment trial carries,
but not for a much larger field. `solver = "lowrank"` fits a Deterministic
Training Conditional (DTC) sparse Gaussian process (Quinonero-Candela and
Rasmussen 2005) on a set of $m$ space-filling inducing points, at cost
linear in the number of plots $n$ and quadratic in the inducing count.
Writing $K_{nm}$ for the covariance between training points and inducing
points and $K_{mm}$ for the inducing points' own covariance, DTC replaces
the exact $n \times n$ training covariance $K_n$ by the rank-$m$
approximation

$$
K_n \approx K_{nm} K_{mm}^{-1} K_{mn},
$$

so every quantity that would otherwise need an $n \times n$ Cholesky
factorisation instead needs only an $m \times m$ one. Every verb above --
point and block prediction, change of support, abstention, manifest
emission -- is unchanged; only the backend differs. This section needs no
real trial data, so it runs regardless of `agridat`: a synthetic field,
labelled as such, of two smooth sinusoidal gradients plus noise.

```{r lowrank}
set.seed(1L)
big <- expand.grid(x = seq(0, 25, by = 0.5), y = seq(0, 25, by = 0.5))
big$yield <- sin(0.3 * big$x) + cos(0.25 * big$y) + rnorm(nrow(big), 0, 0.2)
nrow(big)

fit_lr <- gp_fit(
  gpfield_spec(c("x", "y"), "yield", solver = "lowrank", n_inducing = 120L),
  big, seed = 1L
)
fit_lr
```

When the inducing set covers the whole field the low-rank posterior is
exactly the exact GP, so the choice is a scale-versus-cost dial rather
than a change of model; the default backend stays exact.

## Read

```{r read-prep, echo = FALSE}
if (has_agridat) {
  read_range <- round(fit@hyper$range_raw, 2L)
  read_south_mean <- block_tab$mean[block_tab$block == "south"]
  read_north_mean <- block_tab$mean[block_tab$block == "north"]
  read_diff <- round(read_north_mean - read_south_mean, 1L)
  read_ratio <- round(block_pred@sd[[1L]] / naive_sd, 2L)
  read_coarse_spacing <- round(coarse$diagnostics$spacing, 1L)
  read_coarse_range <- round(coarse$diagnostics$range, 2L)
  read_coarse_ratio <- round(
    coarse$diagnostics$spacing / coarse$diagnostics$range, 1L
  )
  read_far_reach <- round(far_pred$diagnostics$reach, 2L)
  read_verify_ok <- verify_manifest(manifest)$ok
} else {
  read_range <- NA_real_
  read_south_mean <- NA_real_
  read_north_mean <- NA_real_
  read_diff <- NA_real_
  read_ratio <- NA_real_
  read_coarse_spacing <- NA_real_
  read_coarse_range <- NA_real_
  read_coarse_ratio <- NA_real_
  read_far_reach <- NA_real_
  read_verify_ok <- NA
}
```

```{r read-text, echo = FALSE, results = "asis"}
if (has_agridat) {
  cat(sprintf(
    paste0(
      "County C1's estimated correlation range is %.2f plot units, so a ",
      "query spaced further apart than that asks the fit to interpolate ",
      "between plots it no longer treats as related. Splitting the field ",
      "into a south block (rows 1 to 9) and a north block (rows 10 to ",
      "18) gives block-average yields of %.1f in the south and %.1f in ",
      "the north, a difference of %.1f in the direction the row-yield ",
      "correlation in the county-C1 facts table already predicted. The ",
      "south block's change-of-support standard deviation is %.2f times ",
      "what a naive independence assumption would report over the same ",
      "%d quadrature nodes: positive within-block correlation is not ",
      "averaged away, so the paddock-scale uncertainty the table above ",
      "reports is larger than a cell-wise average would suggest.\n\n",
      "Both refusals above are real. The three-point query grid is ",
      "spaced %.1f plot units apart, roughly %.1f times the estimated ",
      "%.2f-unit range, so `gpfield` abstains with reason ",
      "`range_too_short` rather than interpolate between points it ",
      "cannot relate. The far block sits at least %.2f units outside the ",
      "trial footprint at every corner, further than any training ",
      "plot's correlation reach, so `gpfield` abstains with reason ",
      "`support_gap` rather than average zero informative observations ",
      "into a confident block mean. The point-support manifest emitted ",
      "from the same fit verifies: `verify_manifest()$ok` is `%s`.\n"
    ),
    read_range, read_south_mean, read_north_mean, read_diff, read_ratio,
    n_realised, read_coarse_spacing, read_coarse_ratio, read_coarse_range,
    read_far_reach, read_verify_ok
  ))
} else {
  cat("agridat is not installed on this build, so the corn-trial numbers",
      "above were not computed; only the low-rank section, which needs",
      "no trial data, ran.\n")
}
```

## Limits

This vignette shows one axis-aligned rectangular block per call; `gpfield`
does not yet support irregular or polygonal paddock footprints, so a
management block that is not a rectangle needs approximating by one that
is. The low-rank solver demonstrated in the last section is isotropic (a
single length-scale in every direction); an anisotropic low-rank fit is
not available, only the exact solver supports it. Neither block above was
under-supported by a small margin: the south and the north block both sit
comfortably inside the estimated correlation range of every training
plot, and the far block sits far outside it, so the `support_gap`
abstention here demonstrates the guard tripping cleanly rather than the
boundary case where a block sits just inside or just outside the limit.
agridat documents this trial's `yield` column only as "yield", with no
physical unit recorded anywhere in the package, so the figure's colour
scale and every mean and standard deviation above carry Besag and
Higdon's original, unlabelled scale: comparisons within this vignette are
meaningful, a comparison against a value from a different source is not,
unless that source is on the same scale. Finally, nothing here holds out
a plot and checks the fitted surface against it; the fit diagnostics and
the block predictions above describe what the Gaussian process believes,
not how well that belief matches a plot withheld from training.

## What to read next

*The Ensemble-Run Manifest: a Portable, Verifiable Record* (PESTO)
documents the shared contract this vignette's manifest section emits
into, including the reproducibility and integrity guarantees a consumer
can rely on. *From an inference result to a grain plan* (grainPlan) shows
a manifest like this one carried forward into a paddock-level nitrogen,
variety or grade decision, the practical purpose a block-average
prediction like the one above ultimately serves.

## Reproduce

Every fit and prediction above is seeded `seed = 1L`. Block-support
prediction still accepts a `seed` argument for backward compatibility,
but the block quadrature is a deterministic tensor grid and no longer
depends on it. Package versions follow.

```{r session-info}
sessionInfo()
```

## References

Besag, J. and Higdon, D. (1999). Bayesian analysis of agricultural field
experiments. *Journal of the Royal Statistical Society, Series B*, 61(4),
691--746. Trial distributed in the `agridat` package.

Cressie, N. (1993). *Statistics for Spatial Data*, revised edition. Wiley.

Gotway, C. A. and Young, L. J. (2002). Combining incompatible spatial
data. *Journal of the American Statistical Association*, 97(458),
632--648.

Quinonero-Candela, J. and Rasmussen, C. E. (2005). A unifying view of
sparse approximate Gaussian process regression. *Journal of Machine
Learning Research*, 6, 1939--1959.
