---
title: "Change-of-support field smoothing with gpfield"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Change-of-support field smoothing with gpfield}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5
)
set.seed(1L)
```

```{r libs}
library(gpfield)
```

## What this vignette shows

`gpfield` fits a spatial Gaussian process to a field or trial layout and answers
three questions a noisy plot grid alone cannot:

1. what is the smoothed trait surface, with calibrated uncertainty;
2. what is the trait averaged over a paddock-scale block -- a *change of support*
   from the observed point process to an areal one, with the predictive variance
   propagated through the support map rather than averaged away;
3. when is the field simply too coarse, or a block too far from any data, for an
   honest answer -- in which case `gpfield` abstains rather than fabricate a
   confident surface.

We use a real, public multi-environment trial -- the *Besag MET* barley-yield
trial distributed in the `agridat` package -- so every figure below comes from
data you can load yourself. The dataset is documented in `agridat` and originates
with Besag and Higdon (1999).

## A real row-column trial

The trial is laid out as a row-by-column grid of plots within each of six
environments (`county`). We take one environment and treat its plot grid as the
spatial field. Six plots carry a missing yield; `gpfield` drops incomplete rows
before fitting.

```{r data}
if (!requireNamespace("agridat", quietly = TRUE)) {
  knitr::knit_exit()
}
met <- agridat::besag.met
field <- met[met$county == "C1", c("row", "col", "yield")]
str(field)
sum(is.na(field$yield))
```

## Fitting the Gaussian process

A model is declared with `gpfield_spec()` -- which covariance kernel, which
coordinate columns, which response -- and fitted with `gp_fit()`. The default is
a `Matern` covariance at smoothness `1.5`, the usual choice for field data: it is
once-differentiable rather than the squared-exponential's infinite smoothness,
which avoids over-smoothing real agronomic gradients.

```{r fit}
spec <- gpfield_spec(coords = c("row", "col"), response = "yield")
fit <- gp_fit(spec, field, seed = 1L)
fit
```

The fit reports an estimated correlation `range` in raw plot units. That range is
the quantity the rest of the package reasons about: a prediction finer than the
range is unsupported, and the fit makes that explicit instead of leaving it
implicit.

## Smoothing the surface

`gp_field_smooth()` is the one-call path from a row-column layout to a smoothed
surface. Here we refine the prediction grid to twice the plot resolution and draw
the result (the plot needs `ggplot2`).

```{r smooth, fig.alt = "Smoothed barley-yield surface over the row-column plot grid"}
smoothed <- gp_field_smooth(field, response = "yield", refine = 2L, seed = 1L)
smoothed@prediction

if (requireNamespace("ggplot2", quietly = TRUE)) {
  gp_field_plot(smoothed)
}
```

## Change of support: from plots to blocks

The defining capability is predicting at a *different support*. A point process
(the plots) can be aggregated to an areal one (paddock-scale blocks). The block
average is the integral of the latent field over the block, and -- crucially --
its predictive variance is the full quadratic form over the within-block
covariance, so spatial correlation within the block is carried into the
block-average uncertainty.

```{r blocks}
blocks <- block_support(
  lower = rbind(c(1, 1), c(10, 1)),
  upper = rbind(c(9, 11), c(18, 11)),
  ids = c("south", "north")
)
block_pred <- gp_predict(fit, support = "block", blocks = blocks, seed = 1L)
block_pred@locations
data.frame(block = block_pred@locations$block,
           mean = round(block_pred@mean, 1),
           sd = round(block_pred@sd, 2))
```

Why this is not a cell-wise average: a naive mean of `n` point predictions would
assume independence and shrink the block-average standard deviation by
`1 / sqrt(n)`. Because neighbouring locations within a block are positively
correlated, the change-of-support variance is materially larger -- the honest
uncertainty for a paddock-scale claim.

```{r cos-compare}
centre <- gp_predict(fit, data.frame(row = 5, col = 6))
naive_sd <- centre@sd / sqrt(blocks$n_quad)
c(change_of_support_sd = round(block_pred@sd[[1L]], 3),
  naive_independence_sd = round(naive_sd, 4))
```

## Honest abstention

`gpfield` refuses a prediction the data cannot support. Two refusals:

```{r abstain-coarse}
# A query grid spaced far more coarsely than the estimated range.
coarse <- gp_predict(fit, data.frame(row = c(1, 50, 100),
                                     col = c(1, 50, 100)))
is_gpfield_abstention(coarse)
coarse
```

```{r abstain-gap}
# A block placed far outside the trial footprint.
far <- block_support(lower = rbind(c(100, 100)), upper = rbind(c(110, 110)))
gp_predict(fit, support = "block", blocks = far, seed = 1L)
```

Each abstention carries a machine-readable `reason` and the numbers behind it, so
a downstream caller -- or the orchestration layer -- can branch on the refusal
rather than catch an exception or, worse, trust a confident wrong value.

## Emitting the orchestra contract

`gpfield` is an orchestra member, so a result is expressible in the federation's
shared manifest contract: a versioned, integrity-hashed object every member emits
or consumes. `as_orchestra_manifest()` adapts a prediction into it, and
`verify_manifest()` confirms the payload was not tampered with.

```{r manifest}
point_pred <- gp_predict(fit, field[, c("row", "col")])
manifest <- as_orchestra_manifest(point_pred)
manifest
verify_manifest(manifest)$ok
```

The smoothed surface now composes directly with the orchestra's calibration,
uncertainty-quantification, causal-inference and decision members, with no
special-casing on either side.

## Scaling to large fields

The exact solver is cubic in the number of observations -- comfortable for the
hundreds-to-low-thousands of plots a multi-environment trial carries. For a field
too large for the exact solver, `solver = "lowrank"` fits a Deterministic
Training Conditional sparse Gaussian process on a set of space-filling inducing
points, at cost linear in the number of plots and quadratic in the (much smaller)
inducing count. Every verb above -- point and block prediction, change of
support, honest abstention, manifest emission -- is unchanged; only the backend
differs.

```{r lowrank}
set.seed(1L)
big <- expand.grid(x = seq(0, 25, by = 0.5), y = seq(0, 25, by = 0.5))
big$yield <- sin(0.3 * big$x) + cos(0.25 * big$y) + rnorm(nrow(big), 0, 0.2)
nrow(big)

fit_lr <- gp_fit(
  gpfield_spec(c("x", "y"), "yield", solver = "lowrank", n_inducing = 120L),
  big, seed = 1L
)
fit_lr
```

When the inducing set covers the whole field the low-rank posterior is exactly
the exact GP, so the choice is a scale-versus-cost dial rather than a change of
model. The default backend stays exact.

## Reference

Besag, J. and Higdon, D. (1999). Bayesian analysis of agricultural field
experiments. *Journal of the Royal Statistical Society, Series B*, 61(4),
691--746. Dataset distributed in the `agridat` package.
